---
title: "NBA Continuity Analysis"
author: "Michael Culshaw-Maurer"
date: "February 16, 2018"
output: html_document
---

```{r}
library(rethinking)
library(MASS)
require(rvest)
require(tidyverse)
require(ggrepel)
require(ggthemes)
require(zoo)
```

Setting up some functions, web-scraping some of the data
```{r, cache=T}
# web-scraping data on roster continuity
url <- 'https://www.basketball-reference.com/friv/continuity.html'
css_page <- '#continuity'
continuity <- url %>%
  read_html %>%
  html_nodes(css_page) %>%
  html_table(header = T, fill = T) %>%
  data.frame() %>%
  as_tibble()

highlight_teams <- function(team_list, alpha_others){
  # this function will take a list of teams and an alpha value (pick a low value), and return a table with all teams, their colors, and low alpha values for all teams not on your list
  NBA_colors <- c("#E03A3E", "#008248", "black", "#CE1141", "#1D1160", "#6F2633", "#007DC5", "#5091CD", "#ED174C", "#FFCD34", "#CE1141", "#002D62", "#ED174C", "#552583", "#00285E", "#98002E", "#00471B", "#7AC143", "#002B5C", "#F58426", "#007AC1", "#0B77BD", "#006BB6", "#1D1160", "#E13A3E", "#5A2D81", "#C4CED4", "#CD1141", "#0C2340", "#002B5C")
  alpha_values <- rep(alpha_others,30)
  team_names <- c("ATL", "BOS", "BRK", "CHI", "CHO", "CLE", "DAL", "DEN", "DET", "GSW", "HOU", "IND", "LAC", "LAL", "MEM", "MIA", "MIL", "MIN", "NOP", "NYK", "OKC", "ORL", "PHI", "PHO", "POR", "SAC", "SAS", "TOR", "UTA", "WAS")
  
  color_scheme <- as_tibble(cbind(team_names,NBA_colors,alpha_values))
  color_scheme$alpha_values <- as.numeric(color_scheme$alpha_values)
  color_scheme[color_scheme$team_names %in% team_list,3] <- 1
  color_scheme <- as_tibble(color_scheme)
  return(color_scheme)
}
team_list <- NULL
color_scheme <- highlight_teams(team_list = team_list, 1)

```


```{r}
## need to grep out the % symbol and fill in blanks with NAs
continuity2 <- continuity
continuity2[,-1] <- as.numeric(gsub("\\%","",as.matrix(continuity2[,-1])))
continuity3 <- gather(continuity2, "team", "continuity", 2:31)
```


```{r}
ratings <- read_csv("AllYearsTeamRatings.csv")
team_names <- read_csv("team_names_abbreviations.csv")
ratings$Tm.x <- team_names$Tm.x[match(ratings$Team, team_names$Team)]
```

Cleaning everything up
```{r}
ratings[ratings$Team == "New Jersey Nets", 17] <- "BRK"
ratings[ratings$Team == "New Orleans Hornets", 17] <- "NOP"
ratings[ratings$Team == "Charlotte Bobcats", 17] <- "CHO"
ratings[ratings$Team == "Seattle SuperSonics", 17] <- "OKC"
ratings[ratings$Team == "Washington Bullets", 17] <- "WAS"
ratings[ratings$Team == "Vancouver Grizzlies", 17] <- "MEM"
ratings[ratings$Team == "New Orleans/Oklahoma City Hornets", 17] <- "NOP"

continuity3[continuity3$team == "NJN", 2] <- "BRK"
continuity3[continuity3$team == "CHA", 2] <- "CHO"
continuity3[continuity3$team == "NOH", 2] <- "NOP"

continuity3$Season <- gsub("-.*","", continuity3$Season)
continuity3$Season <- as.numeric(continuity3$Season)
continuity3$Season <- continuity3$Season + 1

cont <- full_join(continuity3,ratings, by = c("team" = "Tm.x", "Season" = "Year"))
```


```{r}
cont %>%
  group_by(team) %>% 
  #summarise(mean_cont = mean(cont)) %>% 
  ggplot(aes(x=team, y=continuity))+
  geom_violin(aes(fill=team))+
  coord_flip()+
  theme_tufte()+
  theme(legend.position="none", text=element_text(family="sans", colour="black"))+
  scale_fill_manual(values=color_scheme$NBA_colors)

cont %>%
  ggplot(aes(x=Season, y=continuity))+
  geom_line(aes(group=team, color=team))+
  theme_tufte()+
  theme(legend.position="none", text=element_text(family="sans", colour="black"))+
  scale_colour_manual(values=color_scheme$NBA_colors)+
  facet_wrap(~team)
```

```{r}
cont_rolling <- cont %>% 
  group_by(team) %>% 
  arrange(Season) %>% 
  mutate(roll_cont = rollapply(data = continuity, width = 3, FUN = mean, fill = NA, align = "right"))
```

```{r}
cont_rolling %>%
  ggplot(aes(x=Season, y=roll_cont))+
  geom_line(aes(group=team, color=team))+
  theme_tufte()+
  theme(legend.position="none", text=element_text(family="sans", colour="black"))+
  scale_colour_manual(values=color_scheme$NBA_colors)+
  facet_wrap(~team)
```

```{r}
p <- cont_rolling %>%
  ggplot(aes(x=Season, y=roll_cont))+
  geom_line(aes(group=team, color=team))+
  theme_tufte()+
  scale_colour_manual(values=color_scheme$NBA_colors)+
theme(legend.position="right", panel.grid = element_blank(), legend.title=element_blank(), text=element_text(family="Roboto Condensed", colour="black"))
ggplotly(p)
```



```{r}
d <- cont
```

```{r, cache=TRUE}
 m1 <- map2stan( 
      alist(
          height ~ dnorm( mu , sigma ) ,
          mu <- a[Subject] + bA[Subject]*age,
          c(a,bA)[Subject] ~ dmvnorm2(c(A,B), sigma_Subject, Rho),
          A ~ dnorm(0,1),
          B ~ dnorm(0,1),
          sigma_Subject ~ dcauchy(0,2),
          Rho ~ dlkjcorr(2),
          sigma ~ dcauchy(0,1)
          ),
  
data=d,
warmup=1000 , iter=5000, chains=2, cores=2)
```

**(a)**

```{r}
precis(m6, depth=2)
post6 <- extract.samples(m6)
dens(post6$a)
dens(post6$bA)
```


```{r}
dens( post6$sigma_Subject)
```

This bimodal distribution of sigma shows that there is a lot more variance in the sigma for intercept, which is the peak around 140, whereas the sigma for slope is much lower and peaked.



**(b)**

```{r}

a1 <- apply( post6$a , 2 , mean )
b1 <- apply( post6$bA , 2 , mean )
# plot both and connect with lines
plot( a1 , b1 , xlab="intercept" , ylab="slope" ,
    pch=16 , col=rangi2 , ylim=c( min(b1)-0.1 , max(b1)+0.1 ) ,
    xlim=c( min(a1)-0.1 , max(a1)+0.1 ) )
points( a1 , b1 , pch=1 )
```

There is generally a positive correlation here: if you are taller, your slope, or expected growth rate over the next year, is also higher. For a new group of boys, this would tell you that you would predict the ALREADY tall boys will also grow more quickly per year.



**(c)**

```{r}
mean.a <- mean(post6$a)
mean.bA <- mean(post6$bA)
sd.a <- sd(post6$a)
sd.bA <- sd(post6$bA)
Rho.6 <- mean(post6$Rho)

S2 <- matrix( c(sd.a^2, sd.a*sd.bA*Rho.6, sd.a*sd.bA*Rho.6, sd.bA^2), nrow=2)
```

```{r}
newBOIS <- mvrnorm(n=10, c(mean.a, mean.bA), S2)
```



Let's plot the new boys on the same plot as in 2b, along with the old boys.

```{r}

a1 <- apply( post6$a , 2 , mean )
b1 <- apply( post6$bA , 2 , mean )
# plot both and connect with lines
plot( a1 , b1 , xlab="intercept" , ylab="slope" ,
    pch=16 , col=rangi2 , ylim=c( min(b1)-0.1 , max(b1)+0.1 ) ,
    xlim=c( min(a1)-0.1 , max(a1)+0.1 ) )
points( a1 , b1 , pch=1 )
points(newBOIS, col = "black" )
```

Looks like the new boys fit in nicely!

Let's plot them as trends:  

```{r}
newBOIS <- as.data.frame(newBOIS)
starting_height <- (newBOIS$V1)
ending_height <- (newBOIS$V1 + newBOIS$V2)
starting_age <- c(0,0,0,0,0,0,0,0,0,0)
ending_age <- c(1,1,1,1,1,1,1,1,1,1)

par(mfrow=c(1, 1))
#Plotting the simulated boys
plot( starting_age , starting_height , xlab="age" , ylab="height" ,
pch=16 , col=rangi2 , xlim=c(0,1)
, ylim=c( 120,180 ))
points( ending_age , ending_height , pch=1 )
for ( i in 1:10 ) lines( c(starting_age[i],ending_age[i]) , c(starting_height[i],ending_height[i]) )
```

Yeah, this seems to really confirm the idea that starting height has much more affect on future variance than slope does.
